---
title: "Lab 5"
output: html_document
date: "2023-11-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## This file is the lab for HPM 881 Unit 5 -- Interpretation
Interpretation of racial categories using 3 different approaches:
1) unadjusted, 2) residual direct effect, 3) IOM approach

## Read in the data
```{r}
library(haven)
lab5 <-read_dta('./lecture5_lab.dta')
```

## 1. Get to know your data
Drop observations with <=0 years of education and with missing/invalid responses to RTHLTH53.
```{r}
data <- subset(lab5, EDUCYR > 0 & RTHLTH53 > 0)
```

 Describe the variables of interest
```{r}
summary(lab5[c("DVTEXP18", "RACEV1X", "AGELAST", "SEX", "INSCOV18", "EDUCYR", "RTHLTH53")])
```

 Save the cleaned data
```{r}
saveRDS(data, "lecture5_lab.rds")
```

## 2. Summarize dental expenditures by race categories in RACEV1X.
Using dplyr package
```{r}
library(dplyr)

data %>%
  group_by(RACEV1X) %>%
  summarise(mean_DVTEXP18 = mean(DVTEXP18, na.rm = TRUE))

```


## 3. Create a bar graph showing the ratio of mean dental expenditures for each race category in RACEV1X relative to White.
 For example, if mean dental expenditures for White people = $150 and mean dental expenditures for Black people = $100, the Black/White ratio would = 0.67.
 
 Using ggplot2 package
```{r}


library(ggplot2)

data %>%
  group_by(RACEV1X) %>%
  summarise(mean_DVTEXP18 = mean(DVTEXP18, na.rm = TRUE)) %>%
  mutate(dvt_ratio = mean_DVTEXP18 / mean_DVTEXP18[RACEV1X == 1]) %>%
  ggplot(aes(x = factor(RACEV1X), y = dvt_ratio)) +
  geom_bar(stat = "identity") +
  labs(title = "Ratio of Dental Expenditures to White Race") +
  scale_x_discrete(labels = c("White", "Black", "AI/AN", "Asian/NH/PI", "Multiple"))

```
 

## 4. Unadjusted means approach: Using t-tests, test for differences in mean dental expenditures between each race category in RACEV1X and White.

Loop over values of RACEV1X
```{r}

race_categories <- c(2, 3, 4, 6)

for (i in race_categories) {
  cat("T-test for race category", i, "vs White:\n")
  t_test_result <- t.test(DVTEXP18 ~ factor(RACEV1X), data = data, subset = RACEV1X %in% c(1, i))
  print(t_test_result)
}
```


## 5. Residual direct effect: Regression DVTEXP18 on RACEV1X, AGELAST, EDUCYR, SEX, INSCOV, and RTHLTH53 using White as the reference category.
Interpret any statistically significant differences in the RACEV1X categories. Be precise in your language.

Using lm function
```{r}

model <- lm(DVTEXP18 ~ factor(RACEV1X) + AGELAST + EDUCYR + factor(SEX) + factor(INSCOV18) + RTHLTH53, data = data)
summary(model)
```


## 6. IOM approach: Which variables would you include in the “Clinical Appropriateness and Need” category?
 Which variables would you include in the “Operation of health care systems and legal and regulatory environment” category?
NOTE: We do not have measures of patient preferences.

 Clinical appropriateness and need: AGELAST SEX RTHLTH53
 Operation: EDUCYR INSCOV
```{r}

clinical_vars <- c("AGELAST", "SEX", "RTHLTH53")
operation_vars <- c("EDUCYR", "INSCOV18")

```

